---
title: "Storm Research"
author: "Ian Wheeler"
date: "`r Sys.Date()`"
output: html_document
---

## Abstract

The following is a basic analysis of severe weather data from the U.S. National
Oceanic and Atmospheric Administration's (NOAA) storm database. This database
contains information from 1950 to 2011 related to the characteristics of major
storms and weather events in the United States, including injuries, fatalities,
and property damage. This analysis addresses the questions of which types of
severe weather events are most harmful with respect to population health and
which types of events have the greatest economic consequences. The researcher has
selected property damage estimates and injuries as the metrics by which to measure
economic impact and harm to population health respectively. 

## Results

The data indicate clearly that tornadoes are the most costly severe weather events
in terms of personal injuries, while floods are the most costly in terms of
estimated property damage.


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE)
# Remove scientific notation from output.
options(scipen = 999, digits = 6)
```

## Data Processing

```{r load libraries and data}
setwd("~/R for DS/StormResearch/")

sessionInfo()

library(tidyverse)
library(ggplot2)
library(httr)
library(readr)

# Set source url
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# Save local file destination
src <- "src/StormData.bz2"

# Download the file
download.file(url, destfile = src, method = "curl")

# Read the .bzip2 file into a data frame
df <- read.csv(pipe(paste("bzcat", src)), header = TRUE)

```

## Explore Data

```{r explore data}
head(df,10)
summary(df)
```
## Clean up formatting and filter data

Because the earlier years represented in the database record fewer events
and are less reliable due to inferior record-keeping, this analysis focuses
on the 25 year period from 1986 to 2011.

```{r munge property damage estimates}
beg <- ymd("1986-01-01", tz = "America/Chicago")
end <- ymd("2011-11-30", tz = "America/Chicago")

# Convert BGD_DATE field from character to date; Create normalized property
# damage field by using PROPDMG value multiplied by the PROPDMGEXP value. The
# property damage is normalized to represent millions of dollars.
df <- df %>% 
        mutate(DMGCOSTNORM = case_when(
                PROPDMGEXP == "M" ~ (PROPDMG * 1000000),
                PROPDMGEXP == "B" ~ (PROPDMG * 1000000000),
                PROPDMG > 0 ~ (PROPDMG * 1000)),
               DMGCOSTNORM = ifelse(DMGCOSTNORM > 0, DMGCOSTNORM / 1000000,
                                    DMGCOSTNORM),
               BGN_DATE = mdy_hms(BGN_DATE, tz = "America/Chicago")) %>% 
        filter(BGN_DATE >= beg, BGN_DATE <= end) %>% 
        relocate(DMGCOSTNORM, .before = CROPDMG)
```


```{r injury analysis}
# Analyze Event Types by Injury count
injuries <- df %>%
  group_by(EVTYPE) %>%
  summarize(totalInjuries = sum(INJURIES, na.rm = TRUE)) %>% 
        filter(totalInjuries > 50) %>%
  arrange(desc(totalInjuries))

# Display top 5 injurious events
head(injuries, 5)

# Create a bar plot showing injuries counts by events with more than 50 injuries
ggplot(injuries, aes(x = reorder(EVTYPE, -totalInjuries), y = totalInjuries)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Total Injuries by Event Type",
       x = "Event Type",
       y = "Total Injuries") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Calculate total injuries from severe weather events and total minus tornadoes.
allInj <- sum(df$INJURIES)
twisterInj <- df %>% 
        filter(EVTYPE == "TORNADO") %>% 
        summarize(twistInj = sum(INJURIES))
allTornado <- twisterInj[1,1]
allOthers <- allInj - allTornado
```

The data in this plot are difficult to analyze because of the extremely high
number of tornado injuries as compared to injuries from other events.
tornadoes accounted roughly three-fifths of all severe weather injuries throughout
the United States from 1986 to 2011. During this period, there were `r allInj`
total severe weather injuries, of which `r allTornado` were from tornadoes and
`r allOthers` were from other events.

```{r plot without tornadoes}
# Plot non-tornado injuries
nonTornado <- df %>%
  group_by(EVTYPE) %>%
  summarize(totalInjuries = sum(INJURIES, na.rm = TRUE)) %>% 
        filter(totalInjuries > 50,
               EVTYPE != "TORNADO") %>%
  arrange(desc(totalInjuries))

# Create a bar plot showing injuries counts by non-Tornado events with more than 
# 50 injuries
ggplot(nonTornado, aes(x = reorder(EVTYPE, -totalInjuries), y = totalInjuries)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Total Injuries by Event Type Excluding tornadoes",
       x = "Event Type",
       y = "Total Injuries") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## Property Damage

```{r property damage analysis}
# Analyze Event Types by Injury count
damage <- df %>%
  group_by(EVTYPE) %>%
  summarize(totalDamage = sum(DMGCOSTNORM, na.rm = TRUE)) %>% 
        filter(totalDamage > 500) %>%
  arrange(desc(totalDamage))

# Display the 5 most costly events in terms of property damage estimates
head(damage, 5)

# Create a bar plot showing injuries counts by events with more than 50 injuries
ggplot(damage, aes(x = reorder(EVTYPE, -totalDamage), y = totalDamage)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Total Property Damage Estimate by Event Type",
       x = "Event Type",
       y = "Total Damage in Millions $") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

